---
title: "Cycle_data"
author: "M.MA"
date: "02/02/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing and calling Libraries
```{r}
#install.packages("tidyverse")
#install.packages("sqldf")
#install.packages("lubridate")
#install.packages("ggplot2")

## helps wrangle data
library("tidyverse")  
## helps wrangle data
library("lubridate") 
##helps visualize data
library("ggplot2") 
##helps with using SQL
library("sqldf")
##displays your working directory
getwd() 
```
## Setting up working directory, and creating dataframes for each .csv file.
```{r}
##Setting Directory
#setwd("Desktop/RLearning/Cycle") 

#COLLECT DATA
#=====================
# Upload Divvy datasets (csv files) here
m1_2021 <- read_csv("202101-divvy-tripdata.csv")
m2_2021 <- read_csv("202102-divvy-tripdata.csv")
m3_2021 <- read_csv("202103-divvy-tripdata.csv")
m4_2021 <- read_csv("202104-divvy-tripdata.csv")
m5_2021 <- read_csv("202105-divvy-tripdata.csv")
m6_2021 <- read_csv("202106-divvy-tripdata.csv")
m7_2021 <- read_csv("202107-divvy-tripdata.csv")
m8_2021 <- read_csv("202108-divvy-tripdata.csv")
m9_2021 <- read_csv("202109-divvy-tripdata.csv")
m10_2021 <- read_csv("202110-divvy-tripdata.csv")
m11_2021 <- read_csv("202111-divvy-tripdata.csv")
m12_2021 <- read_csv("202112-divvy-tripdata.csv")

```

## Inspect the dataframes and look for incongruencies
```{r}
str(m1_2021)
str(m2_2021)
str(m3_2021)
str(m4_2021)
str(m5_2021)
str(m6_2021)
str(m7_2021)
str(m8_2021)
str(m9_2021)
str(m10_2021)
str(m11_2021)
str(m12_2021)
```

## Converting ride_id and rideable_type to character so that they can stack correctly
```{r}
m1_2021 <-  mutate(m1_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type)) 
m2_2021 <-  mutate(m2_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m3_2021<-  mutate(m3_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m4_2021 <-  mutate(m4_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m5_2021 <-  mutate(m5_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m6_2021 <-  mutate(m6_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m7_2021 <-  mutate(m7_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m8_2021 <-  mutate(m8_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m9_2021 <-  mutate(m9_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m10_2021 <-  mutate(m10_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m11_2021 <-  mutate(m11_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
m12_2021<-  mutate(m12_2021, ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
```
```{r}
#Stack individual quarter's data frames into one big data frame
cycle_trips <- bind_rows(m1_2021, m2_2021, m3_2021, m4_2021, m5_2021, m6_2021, m7_2021, m8_2021, m9_2021, m10_2021, m11_2021, m12_2021)

# Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020
cycle_trips <- cycle_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

## CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
```{r}
# Inspect the new table that has been created
colnames(cycle_trips)  #List of column names
nrow(cycle_trips)  #How many rows are in data frame?
dim(cycle_trips)  #Dimensions of the data frame?
head(cycle_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(cycle_trips)  #See list of columns and data types (numeric, character, etc)
summary(cycle_trips)  #Statistical summary of data. Mainly for numerics

# Begin by seeing how many observations fall under each usertype
table(cycle_trips$member_casual)
# Add columns that list the date, month, day, and year of each ride
# This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level
# https://www.statmethods.net/input/dates.html more on date formats in R found at that link
cycle_trips$date <- as.Date(cycle_trips$started_at) #The default format is yyyy-mm-dd
cycle_trips$month <- format(as.Date(cycle_trips$date), "%m")
cycle_trips$day <- format(as.Date(cycle_trips$date), "%d")
cycle_trips$year <- format(as.Date(cycle_trips$date), "%Y")
cycle_trips$day_of_week <- format(as.Date(cycle_trips$date), "%A")

# Add a "ride_length" calculation to all_trips (in seconds)
# https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html
cycle_trips$ride_lengths <- difftime(cycle_trips$ended_at,cycle_trips$started_at)

# Inspect the structure of the columns
str(cycle_trips)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(cycle_trips$ride_lengths)
cycle_trips$ride_lengths <- as.numeric(as.character(cycle_trips$ride_lengths))
is.numeric(cycle_trips$ride_lengths)

# Remove "bad" data
# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
# We will create a new version of the dataframe (v2) since data is being removed
cycle_trips_v2 <- cycle_trips[!(cycle_trips$ride_lengths<0),]
str(cycle_trips_v2)
```


## CONDUCT DESCRIPTIVE ANALYSIS

```{r}
# Descriptive analysis on ride_length (all figures in seconds)
mean(cycle_trips_v2$ride_lengths) #straight average (total ride length / rides)
median(cycle_trips_v2$ride_lengths) #midpoint number in the ascending array of ride lengths
max(cycle_trips_v2$ride_lengths) #longest ride
min(cycle_trips_v2$ride_lengths) #shortest ride

# You can condense the four lines above to one line using summary() on the specific attribute
summary(cycle_trips_v2$ride_lengths)

# Compare members and casual users
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual, FUN = mean)
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual, FUN = median)
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual, FUN = max)
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual, FUN = min)


# See the average ride time by each day for members vs casual users
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual + cycle_trips_v2$day_of_week, FUN = mean)

# Notice that the days of the week are out of order. Let's fix that.
cycle_trips_v2$day_of_week <- ordered(cycle_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual + cycle_trips_v2$day_of_week, FUN = mean)

# analyze ridership data by type and weekday
cycle_trips_v2%>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_lengths)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts

# Let's visualize the number of rides by rider type
number_rides_type_plot <- cycle_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_lengths)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  scale_fill_manual(values=c(member = 'lightblue',  casual= 'pink')) +
  geom_col(position = "dodge")
number_rides_type_plot
# Let's create a visualization for average duration
avg_duration_plot <- cycle_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_lengths)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  scale_fill_manual(values=c(member = 'lightblue',  casual= 'pink')) +
  geom_col(position = "dodge")
avg_duration_plot 

### Changing the names of the rideable type to remove the underscore
cycle_trips_v2$rideable_type[cycle_trips_v2$rideable_type == "classic_bike"] <- "Classic Bike"
cycle_trips_v2$rideable_type[cycle_trips_v2$rideable_type == "docked_bike"] <- "Docked Bike"
cycle_trips_v2$rideable_type[cycle_trips_v2$rideable_type == "electric_bike"] <- "Electric Bike"

ride_type <- sqldf("select rideable_type,member_casual,count(rideable_type) as num_uses
                    from cycle_trips_v2
                    group by member_casual, rideable_type
                    ORDER BY count(rideable_type) DESC", method = "auto" )


#ride_type

# Let's create a visualization for number of ride_types of cycles between casual and members

bike_plot <- ggplot(data = ride_type, aes(x = rideable_type, y = num_uses, fill = member_casual)) +
  scale_fill_manual(values=c(member = 'lightblue',  casual= 'pink')) +
  geom_col(position = "dodge") +
  geom_text(aes(label = num_uses),  vjust = -0.3 ,colour = "black", 
            position = position_dodge(.9)) +
  labs(title = "Yearly Total Rides Per Rideable Type.", x = "Rideable Type",
       y = "Total Rides", fill = "Rider Type") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
bike_plot
```

## EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
```{r}
counts <- aggregate(cycle_trips_v2$ride_lengths ~ cycle_trips_v2$member_casual + cycle_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = '~/Desktop/RLearning/Cycle/avg_ride_length.csv')
```

